### 2020-06-22 sections/abstract.tex
implements brooke's suggested changes

% !TEX root = ../main.tex
Spiral structure is endemic in the Universe, and the pitch angle of arms in spiral galaxies is an important observable in efforts to discriminate between different mechanisms of spiral arm evolution. In this paper we present a hierarchical Bayesian approach to galaxy pitch angle determination, using spiral arm data obtained through the \textit{Galaxy Builder} citizen science project {\Lingard}. This approach better deals with the large variations in pitch angle between arms in a galaxy than existing methods, and obtains full posterior distributions on parameters. We utilise fit pitch angles to examine the links to morphology reported in other work, finding no link in our data, and examine a simple model of spiral winding, with which our results are consistent.

>>>>

% !TEX root = ../main.tex
Spiral structure is endemic in the Universe, and the pitch angle of arms in spiral galaxies is an important observable in efforts to discriminate between different mechanisms of spiral arm evolution. In this paper we present a hierarchical Bayesian approach to galaxy pitch angle determination, using spiral arm data obtained through the \textit{Galaxy Builder} citizen science project \citep{2020arXiv200610450L}. This approach better deals with the large variations in pitch angle between arms in a galaxy than existing methods, and obtains full posterior distributions on parameters. We utilise fit pitch angles to examine links between bulge and bar strength and pitch angle reported in other work, finding no link in our data. We similarly test the simple model of spiral winding proposed by \citet{2019arXiv190910291P}, which assumes the uniformity of the cotangent of pitch angle between some unknown upper and lower limits, with results supporting this model of transient and reccurent spiral pitch angle.

### 2020-06-22 sections/introduction.tex
implements brooke's suggested changes

Spiral structure is present in a majority of massive galaxies (\citealt{1989gadv.book..151B}, \citealt{2008MNRAS.389.1179L}) yet the formation mechanisms through which spiral structure originates are still hotly debated. Spirals are as diverse as the theories proposed to govern their evolution; from the quintessential pair of well-defined arcs of the grand design spiral, to the fragmented arm segments of the flocculent spiral, to the disjointed multi-armed spiral. These variatons on structure account for $18\%$, $50\%$ and $32\%$ of the population respectively (\citealt{2011ApJ...737...32E}, \citealt{2015yCat..22170032B}, examples of each type are shown in Figure \ref{fig:spiral-galaxy-types}). The Hubble classification scheme \citep{1926ApJ....64..321H}, and its revisions and expansions (\citealt{1961hag..book.....S}, \citealt{1991rc3..book.....D}) contain detailed variations of different types of spiral galaxy, divided by the presence of a bar and ordered by how obvious spiral arm patterns are, how tightly they are wound and the prominence of a central bulge.

\begin{figure*}
  \includegraphics[width=13cm]{plots/galaxy_types.jpg}
  \caption{Examples of the different types of spiral galaxy present in the sky. The left column shows grand design spirals, the middle shows many-armed spirals and the right shows flocculent spirals. Images were taken with the Sloan Digital Sky Survey Telescope.}
  \label{fig:spiral-galaxy-types}
\end{figure*}

@@

Our current understanding of the mechanisms which drive spiral growth and evolution suggest that each of the different forms of spiral galaxy may be triggered primarily by different processes. Grand Design spirals are thought to have undergone a tidal interaction, be driven by a bar (as seen in gas simulations, \citealt{1976ApJ...209...53S}, \citealt{2008A&A...489..115R}, and suggested for stars by Manifold theory, \citealt{2006A&A...453...39R}, \citealt{2009MNRAS.394...67A}, \citealt{2009MNRAS.400.1706A}), or be obeying (quasi-stationary) density wave theory (QSDW theory), in which spiral arms are slowly evolving, ever-present structures in the disc \citep{1964ApJ...140..646L}. Flocculent spirals are thought to be formed through swing amplification (shearing of small gravitational instabilities in the disc), and be transient and reccurent in nature \citep{1966ApJ...146..810J}.

One of the fundamental assumptions of early work on spiral formation mechanisms (primarily QSDW) was that the disc of a galaxy, if unstable to spiral perturbations, would create a stable, static wave which would exist unchanging for many rotational periods \citep{1964ApJ...140..646L}. The motivation for static waves with small numbers of arms (with a preference for $m=2$) was primarily observational; most galaxies show spiral structure with low spiral arm numbers, suggesting that spirals exist for a long time or are continually rebuilt.

Many simulations demonstrate that spirals do not maintain a constant pitch angle, and instead wind-up over time due to the differential rotation of the disc \citep{2013ApJ...763...46B}. Recent research suggests that spirals arms are transient in nature, and continually dissapate and re-form \citep{2014PASA...31...35D}. These spirals can be maintained through the same mechanisms that drive QSDW spirals (i.e. WASER, \citealt{1976ApJ...205..363M}, swing amplification, \citealt{1965MNRAS.130..125G}), but do not require the idealistic disc conditions required for the formation and maintenance of a stationary wave. The pitch angles of these transient spiral arms will decrease due to the differential rotation of the disk, with the density of the arm peaking at some critical pitch angle, before dissapating to be reformed.

In this dynamic picture of spiral arms, pitch angle monotonically decreases from a spiral arm's formation to its dissapation. \citet{2019arXiv190910291P} propose a simple test of spiral arm winding, assuming the cotangent of the pitch angle of a spiral arm evolves linearly with time. They found that the distribution of pitch angles of their sample of 86 galaxies was consistent with this prediction, evidence against QSDW theory in favour of the dynamic spirals produced in many simulations.

We aim to test this idea of spiral winding using data from the \textit{Galaxy Builder} citizen science project for the 196 spiral galaxies present in \Lingard. We make use of Bayesian hierarchical modelling to measure galaxy pitch angle from the spiral arm clusters producted by \textit{Galaxy Builder}. This methodology allows us to quantify the differences in pitch angles between arms in a single galaxy, as well as investigate the distribution of pitch angles in the galaxy population and investigate relationships between pitch angle and other galaxy properties such as stellar mass or morphology.

Using Galaxy Zoo 2 data \citep{Willett2013:1308.3496v2} allows the separation of galaxies by the presence and strength of bars and bulges; simulations of gas in barred galaxies often demonstrate that bars can drive long-term spiral evolution \citep{2008A&A...489..115R}, or boost transient spiral structure \citep{2012MNRAS.426..167G}. Manifold theory is one attempt to determine the orbits of stars in bar-driven spiral arms: it proposes that stars in the vicinity of the unstable Lagrangian points at either end of the bar tend to escape along predictable orbits, governed by invariant manifolds. One of the primary factors influencing the shape of this invariant manifold is the relative strength of the non-axisymmetric forcing caused by the bar, with stronger bars resulting in spirals with larger pitch angles.

Many other systems correlate with spiral morphology, including potential ties to bulge fraction (\citealt{1975A&A....44..363Y}, \citealt{2013MNRAS.436.1074S}, \citealt{2019MNRAS.487.1808M}) and black hole mass (\citealt{2008ApJ...678L..93S}, \citealt{2017MNRAS.471.2187D}, \citealt{2019MS&E..571a2118A}). Stronger bulges and more more massive central black holes have both correlated with more tightly wound spiral arms.

Section \ref{section:spiral_winding} investigates spiral arm winding using the test derived by \cite{2019arXiv190910291P} (uniformity of galaxy pitch angle in $\cot\;\phi$) and concludes using a marginalized Anderson-Darling test that we cannot unilaterally reject winding of this form at the 1\% level. Section \ref{section:morphology_comparision} examines the correlation between pitch angle and bulge size implied by the Hubble sequence, and pitch angle and bar strength implied by Manifold theory, and find no significant correlation.

>>>>

Spiral structure is present in a majority of massive galaxies (\citealt{1989gadv.book..151B}, \citealt{2008MNRAS.389.1179L}) yet the formation mechanisms through which spiral structure originates are still hotly debated. Spirals are as diverse as the theories proposed to govern their evolution; from the quintessential pair of well-defined arcs of the grand design spiral, to the fragmented arm segments of the flocculent spiral, to the disjointed multi-armed spiral. These variations on structure account for $18\%$, $50\%$ and $32\%$ of the population respectively (\citealt{2011ApJ...737...32E}, \citealt{2015yCat..22170032B}, examples of each type are shown in Figure \ref{fig:spiral-galaxy-types}). The Hubble classification scheme \citep{1926ApJ....64..321H}, and its revisions and expansions (\citealt{1961hag..book.....S}, \citealt{1991rc3..book.....D}) contain detailed variations of different types of spiral galaxy, divided by the presence of a bar and ordered by how obvious spiral arm patterns are, how tightly they are wound and the prominence of a central bulge.

\begin{figure*}
  \includegraphics[width=13cm]{plots/galaxy_types.png}
  \caption{Examples of the different types of spiral galaxy present in the sky. The left column shows grand design spirals, the middle shows many-armed spirals and the right shows flocculent spirals. Images were taken with the Sloan Digital Sky Survey Telescope.}
  \label{fig:spiral-galaxy-types}
\end{figure*}

@@

Our current understanding of the mechanisms which drive spiral growth and evolution suggests that each of the different forms of spiral galaxy may be triggered primarily by different processes. Grand Design spirals are thought to have undergone a tidal interaction, be driven by a bar (as seen in gas simulations, \citealt{1976ApJ...209...53S}, \citealt{2008A&A...489..115R}, and suggested for stars by Manifold theory, \citealt{2006A&A...453...39R}, \citealt{2009MNRAS.394...67A}, \citealt{2009MNRAS.400.1706A}), or be obeying (quasi-stationary) density wave theory (QSDW theory), in which spiral arms are slowly evolving, ever-present structures in the disc \citep{1964ApJ...140..646L}. Flocculent spirals are thought to be formed through swing amplification (shearing of small gravitational instabilities in the disc), and be transient and recurrent in nature \citep{1966ApJ...146..810J}.

One of the fundamental assumptions of early work on spiral formation mechanisms (primarily QSDW) was that the disc of a galaxy, if unstable to spiral perturbations, would create a stable, static wave which would exist unchanging for many rotational periods \citep{1964ApJ...140..646L}. The motivation for static waves with small numbers of arms (with a preference for $m=2$) was primarily observational; most disc galaxies show spiral structure with low spiral arm numbers, suggesting that spirals exist for a long time or are continually rebuilt.

Many simulations demonstrate that spirals do not maintain a constant pitch angle, and instead wind-up over time due to the differential rotation of the disc \citep{2013ApJ...763...46B}. Recent research suggests that spirals arms are transient in nature, and continually dissipate and re-form \citep{2014PASA...31...35D}. These spirals can be maintained through the same mechanisms that drive QSDW spirals (i.e. WASER, \citealt{1976ApJ...205..363M}, swing amplification, \citealt{1965MNRAS.130..125G}), but do not require the idealistic disc conditions required for the formation and maintenance of a stationary wave. The pitch angles of these transient spiral arms will decrease due to the differential rotation of the disk, with the density of the arm peaking at some critical pitch angle, before dissipating to be reformed.

In this dynamic picture of spiral arms, pitch angle monotonically decreases from a spiral arm's formation to its dissipation. \citet{2019arXiv190910291P} propose a simple test of spiral arm winding, assuming the cotangent of the pitch angle of a spiral arm evolves linearly with time. They found that the distribution of pitch angles of their sample of 86 galaxies was consistent with this prediction, evidence against QSDW theory in favour of the dynamic spirals produced in many simulations.

We aim to test this idea of spiral winding using data from the \textit{Galaxy Builder} citizen science project for the spiral galaxies present in \citet{2020arXiv200610450L}. We make use of Bayesian hierarchical modelling to measure galaxy pitch angle from the spiral arm clusters produced by \textit{Galaxy Builder}. This methodology allows us to quantify the differences in pitch angles between arms in a single galaxy, as well as investigate the distribution of pitch angles in the galaxy population and investigate relationships between pitch angle and galaxy morphology.

Using Galaxy Zoo 2 data \citep{Willett2013:1308.3496v2} allows the separation of galaxies by the presence and strength of a stellar bar; simulations of gas in barred galaxies often demonstrate that bars can drive long-term spiral evolution \citep{2008A&A...489..115R}, or boost transient spiral structure \citep{2012MNRAS.426..167G}. Manifold theory is one attempt to determine the orbits of stars in bar-driven spiral arms: it proposes that stars in the vicinity of the unstable Lagrangian points at either end of the bar tend to escape along predictable orbits, governed by invariant manifolds. One of the primary factors influencing the shape of this invariant manifold is the relative strength of the non-axisymmetric forcing caused by the bar, with stronger bars resulting in spirals with larger pitch angles.

Many other systems correlate with spiral morphology, including potential ties to bulge fraction (\citealt{1975A&A....44..363Y}, \citealt{2013MNRAS.436.1074S}, \citealt{2019MNRAS.487.1808M}) and black hole mass (\citealt{2008ApJ...678L..93S}, \citealt{2017MNRAS.471.2187D}, \citealt{2019MS&E..571a2118A}). Stronger bulges and more massive central black holes have both correlated with more tightly wound spiral arms.

Section \ref{section:spiral_winding} investigates spiral arm winding using the test derived by \cite{2019arXiv190910291P} (uniformity of galaxy pitch angle in $\cot\;\phi$). Section \ref{section:morphology_comparision} examines the correlation between pitch angle and bulge size implied by the Hubble sequence, and pitch angle and bar strength implied by Manifold theory.

Where necessary, we make use of $H_0 = 70\ \text{km}\ \text{s}^{-1}\ \text{Mpc}^{-1}$.

### 2020-06-22 sections/method.tex
implements brooke's suggested changes

Many methodologies have been proposed and implemented to measure spiral arm properties, including visual inspection (\citealt{2015A&A...582A..86H}), fourier analysis (i.e. \textsc{2DFFT}, \citealt{2012ApJS..199...33D}), texture analysis (i.e. SpArcFiRe, \citealt{2014ApJ...790...87D}), and combinations of automated methods and human classifiers (\citealt{2017MNRAS.472.2263H}, \citealt{2020MNRAS.493.3854H}). One potentially underused method of obtaining measurements of spirals is through photometric fitting of spiral structure, as possible using tools such as \textsc{GALFIT} \citep{2010AJ....139.2097P} and \textit{Galaxy Builder} (\Lingard). These methods attempt to separate light from an image of a galaxy into distinct subcomponents, such as a galaxy disc, bulge, bar and spiral arms, generally finding the optimum solution using computational optimization. This optimization process, however, is often not robust for complex, many-component models and requires significant supervision to converge to a physically meaningful result \citep{Gao2017:1709.00746v1}. \Lingard propose a solution to this problem through the use of citizen science to provide priors on parameters used in computational fitting.

@@

\begin{equation}
r = A\,e^{\theta\tan\phi},
\end{equation}

where $\phi$ is the arm's pitch angle. One method used to obtain a pitch angle of a galaxy is to fit logarithmic spirals to individually identified arm segments and take the weighted mean of their pitch angles (which often vary by upwards of $10^\circ$, \citealt{2014ApJ...790...87D}). Weighting is determined by the length of the arc segment, with longer being assigned higher weights, i.e. for a galaxy where we have identified $N$ arm segments, each with length $L_i$ and pitch angle $\phi_i$

@@

A notable drawback of length-weighted pitch angle is sensitivity to the number and quality of the spiral arm segments identified; \citet{2017MNRAS.472.2263H} found that only 15\% of the arm segments identified by a leading algorithm (\textsc{SpArcFiRe}) were identified as ``good'' matches to real spiral arms by citizen science classifiers.

Fourier analysis in one- and two-dimensions (as performed by \citealt{2019arXiv190804246D}, \citealt{2012ApJS..199...33D}, \citealt{2018MNRAS.474.2594M}) is another widely used method of computationally obtaining galaxy pitch angles. Two-dimensional fourier methods generally decompose a deprojected image of a galaxy into a superpositions of logarithmic spirals between inner and outer annuli \citep{2012ApJS..199...33D}, and reports the pitch angle with the highest amplitude as the galaxy's pitch angle. \citet{2020MNRAS.493.3854H} combined fourier analysis of spiral galaxies with visual tracing of spiral arms, to great effect. It is uncertain how variation between pitch angles of individual arms impacts this measurement.

\subsection{The Galaxy Sample}
The galaxies analysed in this paper are the 198 galaxies from \Lingard. These are a subset of the \textit{stellar mass-complete sample} in \citet{2017MNRAS.472.2263H}, a sample of low-redshift face-on spiral galaxies selected using data from the NASA-Sloan Atlas \citep{2011AJ....142...31B} and Galaxy Zoo 2 \citep{Willett2013:1308.3496v2}.

Some galaxies in \Lingard were shown to volunteers a second time in a repeat \textit{validation subset} in order to create a second aggregate model used to test internal consistency. We combine the classifications of galaxies in this \textit{validation subset} with the original classifications. Clustering of drawn spiral arms and cleaning of points was then performed as detailed in \Lingard. We remove any galaxies for which no spiral arms were identified, resulting in a hierarchical data structure of 139 galaxies, 261 spiral arms and 239,947 points.

As we wish to utilize the logarithmic spiral's desirable properties of a constant pitch angle and small number of fittable parameters, we make use of it here without an explicit comparison to other models; a simple visual inspection of the fitted logarithmic spirals suggests that it is an appropriate model. A comparison of a logarithmic spiral profile to other spiral forms (i.e. Archimedian or polynomial) is another important piece of work, outside of the scope of this research, as it has been reported that galaxy arms do not have constant pitch angles (\citealt{1981AJ.....86.1847K}; \citealt{2009MNRAS.397..164R}).

Spiral arm points are deprojected to a face-on orientation using the disk inclination and position angle obtained through photometric model fitting performed in \Lingard. Arms are individually corrected to all have the same chirality (a pitch angle greater than or equal to zero) using the logarithmic spiral fit in \Lingard.

@@

In this section we lay out our bayesian hierarchical model for galaxy pitch angle. We fit directly to clustered, cleaned points from polylines drawn in \textit{Galaxy Builder}, deprojected and unwrapped to polar coordinates. We fit a logarithmic spiral to each clustered arm, with the pitch angles of multiple arms in a single galaxy being drawn from a single parent distribution.

@@

The inverse gamma distribution is used to aid the convergence of the Hamiltonian Monte Carlo (HMC) algorithm used (discussed later). The Cauchy distirbution is equivalent to the Student's t-distribution with one degree of freedom, and was chosen due to its fatter tails than the normal distribution. Our likelihood function for $N$ arms, each of which with $n_\mathrm{arm}$ points is

@@

We assume that the radial error is Gaussian for simplicity of analysis, Shapiro-Wilk tests on the residuals of the logarithmic spirals fit in {\Lingard} suggest that this is not a good assumption, and a more robust likelihood (such as the Student's t-distirbution) would possibly more appropriate.

To perform inference, we make use of the No-U-Turn-Sampler (NUTS, \citealt{2011arXiv1111.4246H}), implemented in PYMC3\footnote{\url{https://docs.pymc.io/}}, an open source probabilistic programming framework written in python \citep{pymc3_paper}. To aid the convergence of MC chains, we scale the radii of deprojected points to have unit variance.

>>>>

Many methodologies have been proposed and implemented to measure spiral arm properties, including visual inspection (\citealt{2015A&A...582A..86H}), Fourier analysis (i.e. \textsc{2DFFT}, \citealt{2012ApJS..199...33D}), texture analysis (i.e. SpArcFiRe, \citealt{2014ApJ...790...87D}), and combinations of automated methods and human classifiers (\citealt{2017MNRAS.472.2263H}, \citealt{2020MNRAS.493.3854H}). One potentially underused method of obtaining measurements of spirals is through photometric fitting of spiral structure, as possible using tools such as \textsc{GALFIT} \citep{2010AJ....139.2097P} and \textit{Galaxy Builder} \citep{2020arXiv200610450L}. These methods attempt to separate light from an image of a galaxy into distinct subcomponents, such as a galaxy disc, bulge, bar and spiral arms, generally finding the optimum solution using computational optimization. This optimization process, however, is often not robust for complex, many-component models and requires significant supervision to converge to a physically meaningful result \citep{Gao2017:1709.00746v1}. \citet{2020arXiv200610450L} propose a solution to this problem through the use of citizen science to provide priors on parameters used in computational fitting.

@@

\begin{equation}
r = A\,e^{\theta\tan\phi},
\end{equation}
%
where $\phi$ is the arm's pitch angle. One method used to obtain a pitch angle of a galaxy is to fit logarithmic spirals to individually identified arm segments and take the weighted mean of their pitch angles (which often vary by upwards of $10^\circ$, \citealt{2014ApJ...790...87D}). Weighting is determined by the length of the arc segment, with longer being assigned higher weights, i.e. for a galaxy where we have identified $N$ arm segments, each with length $L_i$ and pitch angle $\phi_i$

@@

A notable drawback of length-weighted pitch angle is sensitivity to the number and quality of the spiral arm segments identified; \citet{2017MNRAS.472.2263H} found that only 15\% of the arm segments identified using \textsc{SpArcFiRe} were identified as ``good'' matches to real spiral arms by citizen science classifiers.

Fourier analysis in one- and two-dimensions (as performed by \citealt{2019arXiv190804246D}, \citealt{2012ApJS..199...33D}, \citealt{2018MNRAS.474.2594M}) is another widely used method of computationally obtaining galaxy pitch angles. Two-dimensional Fourier methods generally decompose a deprojected image of a galaxy into a superposition of logarithmic spirals between inner and outer annuli \citep{2012ApJS..199...33D} and reports the pitch angle with the highest amplitude as the galaxy's pitch angle. \citet{2020MNRAS.493.3854H} combined Fourier analysis of spiral galaxies with visual tracing of spiral arms, to great effect. It is uncertain how the variation between pitch angles of individual arms impacts this measurement.

\subsection{The Galaxy Sample}
The galaxies analysed in this paper are those for which photometric models were obtained in \citet{2020arXiv200610450L}. These are a subset of the \textit{stellar mass-complete sample} in \citet{2017MNRAS.472.2263H}, a sample of low-redshift face-on spiral galaxies selected using data from the NASA-Sloan Atlas \citep{2011AJ....142...31B} and Galaxy Zoo 2 \citep{Willett2013:1308.3496v2}.

Some galaxies in \citet{2020arXiv200610450L} were shown to volunteers a second time in a repeat \textit{validation subset} in order to create a second aggregate model used to test internal consistency. We combine the classifications of galaxies in this \textit{validation subset} with the original classifications. Clustering of drawn spiral arms and cleaning of points was then performed as detailed in \citet{2020arXiv200610450L}. We remove any galaxies for which no spiral arms were identified, resulting in a hierarchical data structure of 139 galaxies, 261 spiral arms and 239,947 points.

As we wish to utilize the logarithmic spiral's desirable properties of a constant pitch angle and a small number of free parameters, we make use of it here without an explicit comparison to other models; a simple visual inspection of the fitted logarithmic spirals suggests that it is an appropriate model. A comparison of a logarithmic spiral profile to other spiral forms (i.e. Archimedian or polynomial) is another important piece of work, outside of the scope of this research, as it has been reported that galaxy arms do not have constant pitch angles (\citealt{1981AJ.....86.1847K}; \citealt{2009MNRAS.397..164R}).

Spiral arm points are deprojected to a face-on orientation using the disk inclination and position angle obtained through photometric model fitting performed in \citet{2020arXiv200610450L}. Arms are individually corrected to all have the same chirality (a pitch angle greater than or equal to zero) using the logarithmic spiral fit in \citet{2020arXiv200610450L}.

@@

In this Section, we lay out our bayesian hierarchical model for galaxy pitch angle. We fit directly to clustered, cleaned points from polylines drawn in \textit{Galaxy Builder}, deprojected and unwrapped to polar coordinates. We fit a logarithmic spiral to each clustered arm, with the pitch angles of multiple arms in a single galaxy being drawn from a single parent distribution.

@@

The inverse gamma distribution is used to aid the convergence of the Hamiltonian Monte Carlo (HMC) algorithm used (discussed later). The Cauchy distribution is equivalent to the Student's t-distribution with one degree of freedom, and was chosen due to its fatter tails than the normal distribution. Our likelihood function for $N$ arms, each with $n_\mathrm{arm}$ points, is

@@

We assume that the radial error is Gaussian for simplicity of analysis, however, Shapiro-Wilk tests on the residuals of the logarithmic spirals fit in \citet{2020arXiv200610450L} suggest that this is not a good assumption, and a more robust likelihood (such as the Student's t-distribution) would possibly more appropriate.

To perform inference, we make use of the No-U-Turn-Sampler (NUTS, \citealt{2011arXiv1111.4246H}), implemented in PYMC3\footnote{\url{https://docs.pymc.io/}}, an open-source probabilistic programming framework written in Python \citep{pymc3_paper}. To aid the convergence of MC chains, we scale the radii of deprojected points to have unit variance.
